---
title: "Class11 HW"
author: "Isabella Ruud PID A59016138"
format: pdf
---

>Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes. 

Hint: The read.table(), 
summary() and boxplot() functions will likely be useful here. There is an example R script online to be used ONLY if you are struggling in vein.  Note that you can find the medium value from saving the output of the boxplot() function to an R object and examining this object. There is also the medium() and summary() function that you can use to check your understanding. 

```{r}
#read in txt file 
expr <- read.table("sample geno exp.txt")

#look at file to make sure it looks okay
head(expr)

#get the sample size for each genotype
table(expr$geno)
```

There are 108 A/A genotypes, 233 A/G genotypes, and 121 G/G genotypes.


```{r}
#get the median expression values for each genotype
aa_median <- median(expr[expr$geno == "A/A",]$exp)
ag_median <- median(expr[expr$geno == "A/G",]$exp)
gg_median <- median(expr[expr$geno == "G/G",]$exp)

cat("A/A expression median: ", aa_median, "\n")
cat("A/G expression median: ", ag_median, "\n")
cat("G/G expression median: ", gg_median, "\n")
```

A/A expression median:  31.24847 
A/G expression median:  25.06486 
G/G expression median:  20.07363 

```{r}
#can also get the summary stats (mean, median, quartiles, etc)
cat("A/A genotype:", "\n")
summary(expr[expr$geno == "A/A",])

cat("A/G genotype:", "\n")
summary(expr[expr$geno == "A/G",])

cat("G/G genotype:", "\n")
summary(expr[expr$geno == "G/G",])

```

>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3? 

Hint: An example boxplot is provided overleaf – yours does not need 
to be as polished as this one.  

```{r}
library("ggplot2")
```


```{r}
#make a box plot of the ORDML3 expression data for each genotype

ggplot(expr) +
  aes(x = geno, y = exp, fill = geno) +
  geom_boxplot(notch = TRUE) +
  geom_jitter(width = 0.2, height = 0, alpha = 0.25) +
  labs(x="Genotype", y = "Expression", title = "SNP Effect on ORDML3 Expression", fill = "Genotype") +
  theme(plot.title = element_text(hjust = 0.5))
```

Yes, the SNP has an effect on the relative expression of ORDML3. The A/A genotype has the highest expression and the G/G genotype has the lowest expression with A/G sitting in between them. It seems that as the number of G alleles increases, the expression of ORDML3 decreases. 














