---
title: "Finishing up Class 10 and Class 12"
author: "Isabella Ruud: PID A59016138"
format: html
---

Let's finish off class 10 with a pipeline for comparative structure analysis. 

```{r}
library(bio3d)
```

```{r}
id <- "1ake_A"
aa <- get.seq(id)
aa
```

lets search for related sequences in the PDB database

```{r}
blast <- blast.pdb(aa)
```

Let's plot an overview of the search results

```{r}
hits <- plot(blast)
```


let's see what the top hits are

```{r}
hits$pdb.id
```

We can download all of these with the 'get.pdb()' function:

```{r}
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

Align the PDBs

```{r}
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```

## Conventional analysis

An RMSD analysis

```{r}
rd <- rmsd(pdbs)
```

```{r}
library(pheatmap)
pheatmap(rd)
```

```{r}
source("https://tinyurl.com/newviewngl")
library(NGLVieweR)
```

```{r}
#view.pdbs(pdbs)
```

We can cluster by RMSD and then color our structure display

```{r}
km <- kmeans(rd, centers=2)
mycols <- c("red", "blue")[km$cluster]
```

```{r}
#view.pdbs(pdbs, color = mycols)
```

## Principal Component analysis

```{r}
pc <- pca(pdbs)
plot(pc, col=mycols)
```

```{r}
p <- mktrj(pc, file = "pca.pdb")
p <- read.pdb("pca.pdb", multi = T)
#view.pdb(p)
```
```{r}
ref <- read.pdb("1ake")
#view.pdb(ref)
```


## AlphaFold Analysis

Here we demonstrate how to analyze and make sense of models from AlphaFold. We begin by reading in all the model PDB files

```{r}
library(bio3d)
files <- list.files("./hiv_monomer_94b5b", pattern = ".pdb", full.names = T)
files
```

align and superpose

```{r}
pbds <- pdbaln(files, fit=TRUE, exefile = "msa")
```

```{r}
source("https://tinyurl.com/newviewngl")
library(NGLVieweR)
#view.pdbs(pdbs)
```



RMSD analysis

```{r}
rd <- rmsd(pdbs)
summary(rd)
```
```{r}
library(pheatmap)

pheatmap(rd)
```



PCA

```{r}
pc <- pca(pdbs)
plot(pc)
```

# Residue conservation from alignment file

AlphaFold writes out the MSA it calculated and used for structure prediction to a A3M format file that we can read into R for further analysis

```{r}
aln_file <- list.files("hiv_monomer_94b5b/", pattern = ".a3m$", full.names = TRUE)
aln_file
```

```{r}
aln <- read.fasta(aln_file, to.upper=TRUE)
```

```{r}
dim(aln$ali)
```

```{r}
sim <- conserv(aln)
```

```{r}
con <- consensus(aln, cutoff = 0.9)
con$seq
```

Plot the conservation along the sequence/structure

```{r}
plot(sim,typ="h")
```

Let's look at these conserved positions

```{r}
pdb <- read.pdb(files[1])
#view.pdb(pdb, backgroundColor="pink", 
         #highlight = atom.select(pdb, resno=25:28),   highlight.style="spacefill") 
```


## HIV dimer

Read in the pdb files that alphafold gave
```{r}
results_dir <- "hiv_dimer_23119.result/hiv_dimer_23119/" 
pdb_files <- list.files(path=results_dir,
                        pattern="*.pdb",
                        full.names = TRUE)

# Print our PDB file names
basename(pdb_files)
```

superimpose the data from the models
```{r}
library(bio3d)

# Read all data from Models 
#  and superpose/fit coords
pdbs <- pdbaln(pdb_files, fit=TRUE, exefile="msa")
```

Find the RMSD between the models
```{r}
rd <- rmsd(pdbs, fit=T)
range(rd)
```

Visualize the RMSD between the models as a heatmap

```{r}
library(pheatmap)

colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```


Plot the pLDDT values across the models

```{r}
plotb3(pdbs$b[1,], typ="l", lwd=2, sse=pdb)
points(pdbs$b[2,], typ="l", col="red")
points(pdbs$b[3,], typ="l", col="blue")
points(pdbs$b[4,], typ="l", col="darkgreen")
points(pdbs$b[5,], typ="l", col="orange")
abline(v=100, col="gray")
```


Improve the superimposition by finding a rigid core 

```{r}
core <- core.find(pdbs)
core.inds <- print(core, vol=0.5)
xyz <- pdbfit(pdbs, core.inds, outpath="corefit_structures")
```


Look at RMSF between positions on the structure

```{r}
rf <- rmsf(xyz)

plotb3(rf, sse=pdb)
abline(v=100, col="gray", ylab="RMSF")
```


Predicted alignment error for domains

```{r}
library(jsonlite)

# Listing of all PAE JSON files
pae_files <- list.files(path=results_dir,
                        pattern=".*model.*\\.json",
                        full.names = TRUE)
```


```{r}
pae1 <- read_json(pae_files[1],simplifyVector = TRUE)
pae5 <- read_json(pae_files[5],simplifyVector = TRUE)

attributes(pae1)

```

```{r}
# Per-residue pLDDT scores 
#  same as B-factor of PDB..
head(pae1$plddt) 
pae1$max_pae
pae5$max_pae
```

```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)")
```

```{r}
plot.dmat(pae5$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

Residue conservation from alignment file

```{r}
aln_file <- list.files(path=results_dir,
                       pattern=".a3m$",
                        full.names = TRUE)
aln_file
```
```{r}
aln <- read.fasta(aln_file[1], to.upper = TRUE)
```

```{r}
dim(aln$ali)
```

```{r}
sim <- conserv(aln)
```

```{r}
plotb3(sim[1:99], sse=trim.pdb(pdb, chain="A"),
       ylab="Conservation Score")
```



```{r}
con <- consensus(aln, cutoff = 0.9)
con$seq
```

```{r}
m1.pdb <- read.pdb(pdb_files[1])
occ <- vec2resno(c(sim[1:99], sim[1:99]), m1.pdb$atom$resno)
write.pdb(m1.pdb, o=occ, file="m1_conserv.pdb")
```









